---
title: "Energy Consumption Analysis"
author: "Nasaka Watitwa"
date: "2024-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Install packages
install.packages("RPostgres")
install.packages("getPass")
# Load library
library(tidyverse)
library(readxl)
library(ggplot2)
library(tidyr)
library(DBI)
library(RPostgres)
library(getPass)
```
# Problem Statement
Analyzing energy generation and consumption in a household over a year from different sources (solar and buying from national grid) to determine profitability of installing a battery for energy storage worth $7000 over buying to supplement solar generation.

# Analysis
```{r}
#Fetch data
energy_2022 <- read_excel("Junior Data Analyst _ Data.xlsx", skip = 2)

#Data structure and missing values
for (c in colnames(energy_2022)) {
  print(paste(c, ":", typeof(energy_2022[[c]]), sum(is.na(energy_2022[[c]]))))
}
```
## Detecting and removing outliers
In order to enhance model performance and accuracy, I made use of the IQR method to detect and remove outliers.
```{r}

boxplot(energy_2022[, c("Solar electricity generation (kWh)", "Electricity usage (kWh)")], 
        main = "Electricity generation and usage",
        names = c(c("Solar electricity generation (kWh)", "Electricity usage (kWh)")),
        ylab = "Values")


is_outlier <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  
    x < lower_bound | x > upper_bound
}


selected_columns <- c("Solar electricity generation (kWh)", "Electricity usage (kWh)")
clean_energy <- energy_2022 %>%
  filter(!(
    is_outlier(.data[[selected_columns[1]]]) |
    is_outlier(.data[[selected_columns[2]]])
  ))

boxplot(clean_energy[, c("Solar electricity generation (kWh)", "Electricity usage (kWh)")],
        main = "Electricity generation and usage",
        names = c(c("Solar electricity generation (kWh)", "Electricity usage (kWh)")),
        ylab = "Values")

```

## Daily trends in hours (without a battery installed)
```{r}
## Hourly energy consumption and generation without battery
hourly_avg <- clean_energy %>% 
  group_by(Hour) %>% 
  summarise(
    generated = pmax(mean(`Solar electricity generation (kWh)`),0),
    consumption = pmax(mean(`Electricity usage (kWh)`),0)
)

ggplot(hourly_avg, aes(x = Hour)) +
  geom_line(aes(y = generated, color = "Solar Electricity Generation (kWh)")) +
  geom_line(aes(y = consumption, color = "Electricity Usage (kWh)")) +
  labs(
    x = "Hour of the Day",
    y = "Energy (KWH)",
    title = "Average Solar Electricity Generation and Electricity Usage per Hour",
    color = "Legend"
  ) +
  theme_minimal()

```


```{r}
##Hourly excess solar electricity generation and purchased electricity
hour_avg <- hourly_avg %>% 
  group_by(Hour) %>% 
  summarise(
    purchased_electricty = mean(ifelse(consumption > generated, consumption - generated, 0), na.rm = TRUE),
    extra_generated = mean(ifelse(generated > consumption, generated - consumption, 0), na.rm = TRUE)
  )

ggplot(hour_avg, aes(x = Hour)) +
  geom_line(aes(y = extra_generated, color = "extra_generated" )) +
  geom_line(aes(y = purchased_electricty, color = "purchased_electricty" )) +
  labs(
    x = "Hour",
    y = "Energy (KWH)",
    title = "Extra solar electricity generation and purchased electricity",
    color = "Legend"
  ) +
  theme_minimal()
```

## Stats with battery installation
```{r}
with_battery <- clean_energy
# Parameters
max_battery_capacity <- 12.5
battery_charge <- 0  # Initial battery charge level

# Initialize columns
with_battery$Battery_Charge_Level <- 0
with_battery$Electricity_Bought <- 0

# Loop through each hour to calculate battery charge and electricity bought
for (i in 1:nrow(with_battery)) {
  solar_generation <- with_battery$`Solar electricity generation (kWh)`[i]
  electricity_usage <- with_battery$`Electricity usage (kWh)`[i]
  if (solar_generation > electricity_usage) {
    excess_generation <- solar_generation - electricity_usage
    charge_to_add <- min(max_battery_capacity - battery_charge, excess_generation)
    battery_charge <- battery_charge + charge_to_add
    with_battery$Electricity_Bought[i] <- 0  # No purchase required
  } else {
    shortfall <- electricity_usage - solar_generation
    discharge_from_battery <- min(battery_charge, shortfall)
    battery_charge <- battery_charge - discharge_from_battery
    with_battery$Electricity_Bought[i] <- shortfall - discharge_from_battery
  }
  with_battery$Battery_Charge_Level[i] <- battery_charge
}
head(with_battery)

```
## Stats without battery installation
```{r}
without_battery <- clean_energy

without_battery <- without_battery %>% 
  mutate(
    Electricity_Bought = ifelse(`Electricity usage (kWh)` > `Solar electricity generation (kWh)`, `Electricity usage (kWh)` - `Solar electricity generation (kWh)`, 0)
  )
```

### Analysis
## Energy savings
```{r}
energy_consumption_comparison <- data.frame(
  Period = c("Electricity Purchased Before Battery", "Electricity Purchased After Battery"),
  Total_Purchased = c(
    sum(without_battery$Electricity_Bought, na.rm = TRUE),
    sum(with_battery$Electricity_Bought, na.rm = TRUE)
  )
)

energy_consumption_comparison$Percentage <- (energy_consumption_comparison$Total_Purchased / energy_consumption_comparison$Total_Purchased[1]) * 100

print(energy_consumption_comparison)

```

## Solar Energy Utilization
```{r}
total_demand <- sum(clean_energy$`Electricity usage (kWh)`)

solar_energy_coverage <- data.frame(
  period = c("Solar utilised before battery", "Solar utilised after battery"),
  Total_Utilised = c(
    sum((without_battery$`Electricity usage (kWh)`- without_battery$Electricity_Bought)),
    sum(with_battery$`Solar electricity generation (kWh)`, na.rm = TRUE)
  )
)

solar_energy_coverage$Coverage_Percentage <- (solar_energy_coverage$Total_Utilised / total_demand) * 100

print(solar_energy_coverage)

```

## Cost Analysis
```{r}
#Grid electricity costs
price_per_kwh <- 22.50
energy_consumption_comparison$Prices <- energy_consumption_comparison$Total_Purchased * price_per_kwh
print(energy_consumption_comparison)

grid_cost_comparison <- data.frame(
  Period = c("Electricity Purchased Before Battery", "Electricity Purchased After Battery"),
  Total_Purchased = c(
    sum(without_battery$Electricity_Bought, na.rm = TRUE),
    sum(with_battery$Electricity_Bought, na.rm = TRUE)
  )
)

grid_cost_comparison <- grid_cost_comparison %>% 
  mutate(Prices = Total_Purchased * price_per_kwh,
         Percentage = (Prices/Prices[1])*100
  )
print(grid_cost_comparison)

#Annual Savings 
annual_savings <- grid_cost_comparison$Prices[1] - grid_cost_comparison$Prices[2]

```


##Projecting 20 years ahead
```{r}
# Inputs
base_savings <- 36257.26
years <- 1:20
discount_rate <- 0.04

# Scenario 1: Fixed 4% increase
scenario1_savings <- base_savings * (1 + 0.04)^(years - 1)

# Scenario 2: Increasing growth rate (4% + 0.25% per year)
scenario2_growth_rates <- 0.04 + 0.0025 * (years - 1)
scenario2_savings <- base_savings * cumprod(1 + scenario2_growth_rates)

# NPV Calculation
scenario1_npv <- sum(scenario1_savings / (1 + discount_rate)^years)
scenario2_npv <- sum(scenario2_savings / (1 + discount_rate)^years)

# Results
cat("NPV for Scenario 1:", "Ksh", scenario1_npv, "\n")
cat("NPV for Scenario 2:", "Ksh", scenario2_npv, "\n")

cat("Svaing for Scenario 1:", "Ksh", sum(scenario1_savings), "\n")
cat("Svaing for Scenario 2:", "Ksh", sum(scenario2_savings), "\n")
```
## Exporting to DB
```{r}
#Database Connection
con <- dbConnect(RPostgres::Postgres(), dbname = "Electricity", user = "postgres", password = getPass::getPass("Database password"))

dbListTables(con)
dbWriteTable(con, "original_data", as.data.frame(energy_2022), overwrite = TRUE, row.names = FALSE)
dbWriteTable(con, "cleaned_data", as.data.frame(energy_cleaned), overwrite = TRUE, row.names = FALSE)
dbWriteTable(con, "with_battery", as.data.frame(with_battery), overwrite = TRUE, row.names = FALSE)
dbWriteTable(con, "without_battery", as.data.frame(without_battery), overwrite = TRUE, row.names = FALSE)


```





